Supplement to Example Manuscript Template for a Data Analysis Project

### 2025-04-18 products/manuscript/Manuscript.qmd
Part 5 of MADA Project

---
title: "Immune-Related Adverse Events (irAEs) associated with Immune Checkpoint Inhibitor (ICI) Therapy in Bladder Cancer: An Analysis of the FDA Adverse Event Reporting System (FAERS)"
author: "Pooja Gokhale, PharmD"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/nlm-style.csl
---

@@

Background: Bladder cancer is the ninth most common type of cancer worldwide. Treatment options include surgery, chemotherapy and immunotherapy. Immune checkpoint inhibitors (ICIs) are a class of immunotherapeutic agents recommended for the treatment of bladder cancer. ICIs have been associated with certain immune-related adverse events (irAEs)

Methods: Reports in the FAERS dashbaord through December 2024 were analyzed for irAEs to four ICIs recommended in bladder cancer - atezolizumab, avelumab, nivolumab, and pembrolizumab. Standardized queries were used to identify cases of rash, pruritus, hypothyroidism, hyperthyroidism, colitis, hepatitis, and nephritis. Descriptive and disproportionality analyses were performed to detect safety signals.

Results: Out of 2948 cases identified from FAERS, 365 cases with irAEs were analyzed. The largest number of cases were reported for pembrolizumab (n=307), followed by atezolizumab (n = 91), avelumab (n = 41), and nivolumab (n =37). Pembrolizumab was found to have safety signals for all irAEs considered together (PRR: 1.8 (1.52, 2.14), ROR: 2.07 (1.69, 2.55), IC:	0.36 (0.29, 0.44)). A safety signal was also found for rash (PRR: 2.11 (1.17, 3.8), ROR:	2.14 (1.18, 3.9), IC:	0.44 (0.17, 0.71)). Safety signals for atezolizumab included hyperthyroidism and nephritis, while hypothyroidism was identified as a concern for avelumab.

@@

Bladder cancer is the ninth most common type of cancer, according to a 2022 report published by the International Agency for Research on Cancer [@Bray2024]. The American Cancer Society estimates approximately 85,000 new bladder cancer cases, and approximately 17,000 bladder cancer deaths in the United States in 2025.

@@

According to the National Comprehensive Cancer Network (NCCN) guidelines, immunotherapy with immune checkpoint inhibitors (ICIs) is recommended at various stages of bladder cancer treatment. ICIs such as pembrolizumab, atezolizumab, nivolumab, and avelumab are FDA approved and recommended for the adjuvant and first and second-line treatment of bladder cancer [@Flaig2024].

ICIs are associated with various adverse effects, particularly with immune-related adverse events (irAEs) such as pneumonitis, hepatitis, colitis, hypothyroidism, dermatitis, and others [@Yin2023].This study aims to analyze real-world adverse event data from the FDA Adverse Event Reporting System associated with ICI therapy in bladder cancer.

@@

The study aims to analyze whether irAEs associated with ICIs generate a signal of disproportionate reporting (SDR). The goal is to determine whether irAEs associated with ICIs in baldder cancer are greater than expected, and thus further research and strategies to deal with these adverse events is required.

@@

A random forest model was fit to distinguish between serious and non-serious cases on the basis of age, sex, and ICI.

@@

## Data aquisition

In order to obtain the FAERS data, the ICIs were used to search by product. Pembrolizumab, atezolizumab, avelumab, and nivolumab were the keywords used. Previously, durvalumab, and ipilimumab were also approved for bladder cancer treatment. However, this indication has since then been removed by the manufacturers. As such, these were not considered. Following this, the search was limited to cases of bladder cancer. A total of 2948 reactions were obtained.

@@

The file was downloaded as an Excel file. First, the column were renamed so as to make them easier to use in code. Then, products with concomitant drugs were removed, so as to focus only on cases associated with ICIs. Then, cases which also had other drugs as the suspect active ingredients were removed. 
This was followed by retaining only the irAEs.
A list of common irAEs was used. This list is available in the supplementary material. Any reactions which had the same age, weight, FDA received date, country of occurrence, suspected active ingredient, and reported reaction were considered duplicates and were removed.
Following this, the reactions were coded according to the MedDRA Preferred Terms (PTs), and classified according to the System Organ Class (SOC).
The age values were converted to numeric and the countries were re-classified as US, outside US, and not specified.
The code for the analysis would involve loading the file, removing duplicates, assessing missing data, and deciding whether to remove it. The details are available in the processing code file.

@@

Firstly, descriptive statistics were used to describe the demographic and other characteristics.
This was followed by creating some descriptive figures.
Two logistic regression models were fit to assess the association by severity with sex and age.
The PRR, ROR and IC (along with their 95% CI) were then calculated to assess the drug signal.
Three different machine learning models were used to distinguish severity on the bases of sex, age, and type of ICI.
Due to missing values, the patient age values were imputed.
First, the data was split into training and testing data (70/30 split). A 5-fold cross-validation was used.

@@

Most of the reaction were reported by healthcare professionals. Since FAERS is a US-based dashboard, many reactions were reported from the US.

@@

Pembrolizumab has the highest number of reactions. The largest percentage of reactions are classified as "Other". These include reactions which do not belong to the other organ classes, such as myositis.

@@

The predicted severity decreases with age.
However, this relationship is also not signficant.

@@

The following tables show the results from the disproportionality analysis.

@@

Similarly, avelumab may has a safety signal for hypothyroidism (PRR: 3.05 (0.98, 9.51), ROR:	3.08 (0.97, 9.75), IC:	1.32 (0.12, 2.52)).

@@

@tbl--randomforest shows the results of the three machine learning models (logistic regression, decision tree, random forest).

@@

# Load and prepare data
rf_metrics <- readRDS("../../results/tables/model_table.rds")

rf_metrics_cleaned <- rf_metrics[-1, ]  # Removes the second row

flextable(rf_metrics_cleaned) |> 
  bold(part = "header") |> 
  border(part = "all", border = fp_border(color = "black", width = 1)) |> 
  colformat_double(digits = 3) |>  
  align(align = "center", part = "header") |> 
  align(align = "right", part = "body", j = which(sapply(rf_metrics_cleaned, is.numeric))) |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  width(width = c(1.8, rep(1.2, ncol(rf_metrics_cleaned)-1))) |>  
  bg(part = "header", bg = "#f2f2f2") |>  
  fontsize(size = 10, part = "all") |> 
  padding(padding = 4, part = "all")

@@

The Random Forest model has an AUC of 0.806. Thus, this models is able to predict between serious and non-serious cases ~81% of the time.
The logistic regression model is next, able to distinguish ~60% of the time.
The decision tree model performs poorly, possibly due to fitting issues.
All the models have the same accuracy. Additionally, sensitivity is 0 and specificity is high, meaning all the models always predict 'serious'.
Such problems arise due to class imbalances, and may require advanced methods to deal with.
The random forest model does not perform as well on the test data (AUC=0.56), and is able to distinguish the severity only 56% of the time.

@@

*More details will be added in part 5 to the discussion section.*

>>>>

---
title: "Immune-Related Adverse Events (irAEs) Associated with Immune Checkpoint Inhibitor (ICI) Therapy in Bladder Cancer: An Analysis of the FDA Adverse Event Reporting System (FAERS)"
author: "Pooja Gokhale, PharmD"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/nlm-style.csl
---

@@

Background: Bladder cancer is the ninth most common type of cancer worldwide. Treatment options include surgery, chemotherapy, and immunotherapy. Immune checkpoint inhibitors (ICIs) are a class of immunotherapeutic agents recommended for the treatment of bladder cancer. ICIs have been associated with certain immune-related adverse events (irAEs)

Methods: Reports in the FAERS dashboard through December 2024 were analyzed for irAEs to four ICIs recommended in bladder cancer: atezolizumab, avelumab, nivolumab, and pembrolizumab. Standardized queries were used to identify cases of rash, pruritus, hypothyroidism, hyperthyroidism, colitis, hepatitis, and nephritis. Descriptive and disproportionality analyses were performed to detect safety signals.

Results: Out of 2948 cases identified from FAERS, 365 cases with irAEs were analyzed. The largest number of cases was reported for pembrolizumab (n=307), followed by atezolizumab (n = 91), avelumab (n = 41), and nivolumab (n =37). Pembrolizumab was found to have safety signals for all irAEs considered together (PRR: 1.8 (1.52, 2.14), ROR: 2.07 (1.69, 2.55), IC:	0.36 (0.29, 0.44)). A safety signal was also found for rash (PRR: 2.11 (1.17, 3.8), ROR:	2.14 (1.18, 3.9), IC:	0.44 (0.17, 0.71)). Safety signals for atezolizumab included hyperthyroidism and nephritis, while hypothyroidism was identified as a concern for avelumab.

@@

Bladder cancer is the ninth most common type of cancer, according to a 2022 report published by the International Agency for Research on Cancer [@Bray2024]. The American Cancer Society estimates approximately 85,000 new bladder cancer cases and approximately 17,000 bladder cancer deaths in the United States in 2025.

@@

According to the National Comprehensive Cancer Network (NCCN) guidelines, immunotherapy with immune checkpoint inhibitors (ICIs) is recommended at various stages of bladder cancer treatment. ICIs such as pembrolizumab, atezolizumab, nivolumab, and avelumab are FDA-approved and recommended for the adjuvant and first and second-line treatment of bladder cancer [@Flaig2024].

ICIs are associated with various adverse effects, particularly with immune-related adverse events (irAEs) such as pneumonitis, hepatitis, colitis, hypothyroidism, dermatitis, and others [@Yin2023]. This study aims to analyze real-world adverse event data from the FDA Adverse Event Reporting System associated with ICI therapy in bladder cancer.

@@

The study aims to analyze whether irAEs associated with ICIs generate a signal of disproportionate reporting (SDR). The goal is to determine whether irAEs associated with ICIs in bladder cancer are greater than expected, and thus, further research and strategies to deal with these adverse events are required.

@@

The association between severity of reactions and age and sex was assessed using logistic regression.
Three machine learning models (logistic regression, decision tree, and random forest) were used to distinguish between serious and non-serious reactions.

@@

## Data acquisition

In order to obtain the FAERS data, the ICIs were used to search by product. Pembrolizumab, atezolizumab, avelumab, and nivolumab were the keywords used. Previously, durvalumab and ipilimumab were also approved for bladder cancer treatment. However, this indication has since then been removed by the manufacturers. As such, these were not considered. Following this, the search was limited to cases of bladder cancer. A total of 2948 reactions were obtained.

@@

The file was downloaded as an Excel file. Then, in R, the columns were renamed so as to make them easier to use in code. Then, products with concomitant drugs were removed so as to focus only on cases associated with ICIs. Then, cases that also had other drugs as the suspect active ingredients were removed. 
This was followed by retaining only the irAEs.
A list of common irAEs was used. This list is available in the supplementary material. This list was informed by common irAEs reported in various clinical trials  [@Galsky2020] [@Galsky2020] [@Bajorin2021] [@Powles2020]. Any reactions that had the same age, weight, FDA received date, country of occurrence, suspected active ingredient, and reported reaction were considered duplicates and were removed.
Following this, the reactions were coded according to the MedDRA Preferred Terms (PTs) and classified according to the System Organ Class (SOC).
The age values were converted to numeric, and the countries were reclassified as US, outside US, and not specified.
The code for the analysis would involve loading the file, removing duplicates, assessing missing data, and deciding whether to remove it. The details are available in the processing code file.

@@

Firstly, descriptive statistics were used to describe the demographic and other characteristics.
This was followed by creating some descriptive figures.
Two logistic regression models were fit to assess the association by severity with sex and age.
The PRR, ROR, and IC (along with their 95% CI) were then calculated to assess the drug signal.
Three different machine learning models were used to distinguish severity on the basis of sex, age, and type of ICI.
Due to missing values, the patient age values were imputed.
First, the data was split into training and testing data (70/30 split). A 5-fold cross-validation was used.

@@

Most of the reactions were reported by healthcare professionals. Since FAERS is a US-based dashboard, many reactions were reported from the US.

@@

Pembrolizumab has the highest number of reactions. The largest percentage of reactions are classified as "Other". These include reactions that do not belong to the other organ classes, such as myositis.

@@

The predicted severity decreases with age.
However, this relationship is also not significant.

@@

The following tables show the results of the disproportionality analysis.

@@

Similarly, avelumab may have a safety signal for hypothyroidism (PRR: 3.05 (0.98, 9.51), ROR:	3.08 (0.97, 9.75), IC:	1.32 (0.12, 2.52)).

@@

@tbl-randomforest shows the results of the three machine learning models (logistic regression, decision tree, random forest).

@@

library(dplyr)
library(flextable)

# Load and clean data
rf_metrics <- readRDS("../../results/tables/model_table.rds")
rf_metrics_cleaned <- rf_metrics[-1, ]

# Convert numeric-looking columns to double (if needed)
rf_metrics_cleaned <- rf_metrics_cleaned %>%
  mutate(across(where(~ all(grepl("^[-+]?[0-9]*\\.?[0-9]+$", .))), as.numeric))

# Build styled flextable
flextable(rf_metrics_cleaned) |> 
  bold(part = "header") |> 
  border(part = "all", border = fp_border(color = "black", width = 1)) |> 
  colformat_double(digits = 3, na_str = "-") |>  
  align(align = "center", part = "header") |> 
  align(
    align = "left",
    part = "body",
    j = which(sapply(rf_metrics_cleaned, is.numeric))
  ) |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  width(width = c(1.8, rep(1.2, ncol(rf_metrics_cleaned) - 1))) |>  
  bg(part = "header", bg = "#f2f2f2") |>  
  fontsize(size = 10, part = "all") |> 
  padding(padding = 4, part = "all")

@@

The Random Forest model has an AUC of 0.806. Thus, this model is able to predict between serious and non-serious cases ~81% of the time.
The logistic regression model is next, able to distinguish ~60% of the time.
The decision tree model performs poorly, possibly due to fitting issues.
All the models have the same accuracy. Additionally, sensitivity is 0, and specificity is high, meaning all the models always predict 'serious'.
Such problems arise due to class imbalances, and may require advanced methods to deal with.
The random forest model does not perform as well on the test data (AUC=0.56), and is able to distinguish the severity only 56% of the time.

@@

### 2025-04-18 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 of MADA Project

This part will be updated in part 5 - after improving the random forest code.
{{< pagebreak >}}

@@

```{r}
#| label: fig-result
#| fig-cap: "Bar chart of outcome of immune checkpoint inhibitors"
#| echo: FALSE
library(here)
library(knitr)
knitr::include_graphics(here("results","figures","figure1.png"))
```

@@

@fig-result2 shows the bar chart by type of organ system involved in the reaction.

```{r}
#| label: fig-result2
#| fig-cap: "ROC Curve - Logistic Regression"
#| echo: FALSE
knitr::include_graphics(here("results","figures","logit_roc.png"))
```

@fig-result3 shows the bar chart by type of organ system involved in the reaction.

@@

@fig-result4 shows the bar chart by type of organ system involved in the reaction.

@@

@@

Additional discussion will be done in part 5 after improving the random effects model.
[@Leslie2025-lp] is a placeholder reference.

# References

>>>>

Three machine learning models were fit - logistic regression, decision tree, and random forest.
We found that the models were not able to distinguish adequately between the serious and non-serious reactions.
This most probably occurs due to the class imbalance - 305 serious, 60 non-serious cases.

@@

```{r}
#| label: fig-result
#| fig-cap: "Bar chart of outcome of immune checkpoint inhibitors"
#| echo: FALSE
library(here)
library(knitr)
knitr::include_graphics(here("results","figures","figure1.png"))
```
We can see that a large proportion of the reactions are serious.

@@

The most common reactions are dermatological, followed by gastrointestinal.

@fig-result2 shows the ROC curve for the logistic regression model.

```{r}
#| label: fig-result2
#| fig-cap: "ROC Curve - Logistic Regression"
#| echo: FALSE
knitr::include_graphics(here("results","figures","logit_roc.png"))
```
We see that the curve covers about 60% of the area.

@fig-result3 shows the ROC curve for the decision tree model.

@@

The curve is a staright diagonal line (covering about 50% of the area). Thus, this model is not able to distinguish between serious and non-serious cases better than at random.

@fig-result4 shows the ROC curve for the random forest model.

@@

We see that the curve covers about 80% of the area.

@@

Overall, the random forest model performs better.
However, all three models are not able to distinguish between serious and non-serious reactions, possibly due to the class imbalance.
